### 2020-03-04 Common/chapters/introduction.tex
Johan and Julien: all names for each section added

According to the World Health Organization, cancer caused $9.6$ million deaths in 2018, making it the second leading cause of death~\cite{44}. An early detection and treatment increases the chances of recovery. In this context, Computer-Aided Diagnosis (CAD) systems can play a massive role by preventing health professionals from missing positive diagnoses. Thanks to the attention that deep learning has gotten over the last decade, newer and better diagnosis systems have transfered the advantages of deep learning to cancer detection, diagnosis and localization tasks.

@@

\section{Motivation}
Cancer is a serious concern in today's society. An early detection is key to maximize the chances of recovery. As explained by Nicholas Petrick~\cite{50}, "it has long been recognized that clinicians do not always make optimal use of the data acquired by an imaging device. The limitations of the human eye-brain system, limitations in training and experience, and factors such as fatigue, distraction, and satisfaction of search may all contribute to suboptimal use of available information". Under these circumstances, CAD systems can contribute to address some of these issues by being used as an aid for clinicians. A complete CAD system is composed of two parts. The first one is called CADe for "computer-aided detection". This part includes medical  image  analysis  tasks  such  as  segmentation,  identification,  localization  and  detection. The other part is CADx for "computer-aided diagnosis" which aims at extracting the characteristics of lesions to classify them according to their malignancy. CAD systems provide multiple advantages.\\
First of all, used as an extra diagnosis, they decrease the probability of missing positive diagnoses. Then, they can also speed up the diagnosis process by proposing regions of interest to clinicians, which, in time, could reduce the screening costs.\\
Similarly, lower screening costs imply a greater accessibility to screening tests, which can result in an earlier cancer detection. As stated before, detecting cancers at an early stage maximizes the probability of curing them. Another benefit of using CAD systems, especially the ones based on deep learning, is the number of cases that the latter are trained on. In fact, in order to built efficient deep learning models, large amounts of data are required. Consequently, efficient CAD systems usually see a lot more cancer cases than beginner clinicians.\\
However, these advantages remain partially theoretical nowadays. Research is still ongoing, aiming at improving existent CAD systems. Problems like poor detection rates, small amounts of available data and bad levels of generalization are still recurrent. Therefore, this situation can and must be improved.

@@

\section{Contributions}
This work proposes a state-of-the-art deep learning model for prostate lesion classification and proposes a method based on transfer learning using different body parts to overcome the lack of publicly available data in the medical field. Concretely, the major contributions of this work are:
\begin{itemize}
\item A state-of-the-art deep learning model for prostate cancer detection. All steps are clearly described, the corresponding scripts are available on GitHub\footnote{\url{https://github.com/eXascaleInfolab/2019_Hospital-Fribourg}} and the raw results are presented. This makes the experiment completely reproductible.

@@

\section{Work repartition}
This thesis was jointly written by two authors. The name(s) of the contributor(s) in the margins of the following chapters only concern the written thesis. Both of them contributed equally to the work leading to the writing of the latter. The work was either done together or in parallel, depending on the tasks. In any case, parts written by one of the authors were reread by the other. Finally, the order in which the names are cited does not represent the quantity of work performed and was chosen arbitrarily.

>>>>

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}According to the World Health Organization, cancer caused $9.6$ million deaths in 2018, making it the second leading cause of death~\cite{44}. An early detection and treatment increases the chances of recovery. In this context, Computer-Aided Diagnosis (CAD) systems can play a massive role by preventing health professionals from missing positive diagnoses. Thanks to the attention that deep learning has gotten over the last decade, newer and better diagnosis systems have transfered the advantages of deep learning to cancer detection, diagnosis and localization tasks.

@@

\section{Motivation}

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Johan}}Cancer is a serious concern in today's society. An early detection is key to maximize the chances of recovery. As explained by Nicholas Petrick~\cite{50}, "it has long been recognized that clinicians do not always make optimal use of the data acquired by an imaging device. The limitations of the human eye-brain system, limitations in training and experience, and factors such as fatigue, distraction, and satisfaction of search may all contribute to suboptimal use of available information". Under these circumstances, CAD systems can contribute to address some of these issues by being used as an aid for clinicians. A complete CAD system is composed of two parts. The first one is called CADe for "computer-aided detection". This part includes medical  image  analysis  tasks  such  as  segmentation,  identification,  localization  and  detection. The other part is CADx for "computer-aided diagnosis" which aims at extracting the characteristics of lesions to classify them according to their malignancy. CAD systems provide multiple advantages.\\
First of all, used as an extra diagnosis, they decrease the probability of missing positive diagnoses. Then, they can also speed up the diagnosis process by proposing regions of interest to clinicians, which, in time, could reduce the screening costs.\\
Similarly, lower screening costs imply a greater accessibility to screening tests, which can result in an earlier cancer detection. As stated before, detecting cancers at an early stage maximizes the probability of curing them. Another benefit of using CAD systems, especially the ones based on deep learning, is the number of cases that the latter are trained on. In fact, in order to built efficient deep learning models, large amounts of data are required. Consequently, efficient CAD systems usually see a lot more cancer cases than beginner clinicians.\\
However, these advantages remain partially theoretical nowadays. Research is still ongoing, aiming at improving existent CAD systems. Problems like poor detection rates, small amounts of available data and bad levels of generalization are still recurrent. Therefore, this situation can and must be improved.

@@

\section{Contributions}

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Johan}}This work proposes a state-of-the-art deep learning model for prostate lesion classification and proposes a method based on transfer learning using different body parts to overcome the lack of publicly available data in the medical field. Concretely, the major contributions of this work are:
\begin{itemize}
\item A state-of-the-art deep learning model for prostate cancer detection. All steps are clearly described, the corresponding scripts are available on GitHub\footnote{\url{https://github.com/eXascaleInfolab/2019_Hospital-Fribourg}} and the raw results are presented. This makes the experiment completely reproductible.

@@

\section{Work repartition}

This thesis was jointly written by two authors. The name(s) of the contributor(s) in the margins of the following chapters only concern the written thesis. Both of them contributed equally to the work leading to the writing of the latter. The work was either done together or in parallel, depending on the tasks. In any case, parts written by one of the authors were reread by the other. Finally, the order in which the names are cited does not represent the quantity of work performed and was chosen arbitrarily.

### 2020-03-04 Common/chapters/literature_review.tex
Johan and Julien: all names for each section added

\chapter{Literature review}
\label{ch:literature_review}
%Traditionally, building a CAD system consisted of two different steps: "Lesion  detection (CADe) and  false-positive  reduction (CADx).  Lesion  detection  was  primarily  based  on  algorithms  specific  to  the  detection task, resulting in many candidate lesions. False-positive reduction was commonly based on traditional  machine  learning  methods  to  reduce  the  false  positive  rate. However,  even  with  these  complicated and sophisticated programs, the general performance of conventional CAD systems was not good, thus hampering their widespread usage in clinical practice"~\cite{41}.\\
%Nowadays, these two steps tend to disappear thanks to deep learning algorithms, in particular the ones based on convolutional neural networks. In fact, the latter can be considered as single-step CAD system. For this reason, CAD systems are currently a trending topic in deep learning.\\
Since this work focuses on lesions classification, this chapter presents the main research works about CADx systems based on convolutional neural networks for each body part used in sections \ref{ch:paper_reproduction} and \ref{ch:transfer_learning}, i.e. prostate, lung and brain. Furthermore, it exclusively focuses on studies that used the same datasets as the ones used in this work.

\section{Prostate/PROSTATEx}
As stated by Gao et al., "Prostate  cancer  is  the  most  common  malignancies  among  men  and  remains  a  second  leading  cause to deaths in men globally. It was predicted that there would be 1.7 million new cancer cases by 2030. The early detection and diagnosis of prostate cancer can help to survive nine out of 10 men for the last five years"\cite{41}. Therefore, researchers proposed many different models to achieve good performance in prostate cancer detection. All the research works mentioned below are based on the SPIE-AAPM-NCI PROSTATEx Challenge dataset. This dataset is composed of multiparametric MRIs (T2W, DWI, ADC, DCE, PD, Ktrans) for a total of 204 training patients and 208 challenge patients (see Section \ref{sec:prostatex_dataset_description}).

@@

\section{Lung/Lung CT Challenge}
Gao et al. claimed that "lung cancer is one of the most frequent and leading causes of death all over the world. It was reported that there were approximately $1.8$x$10^6$ new cases of lung cancer globally in 2012. Early detection of lung cancer, which is typically viewed in the form of lung nodules, is an efficient way to improve the survival rate"~\cite{41}. The articles cited below make use of the SPIE-AAPM Lung CT Challenge dataset, which is composed of CT scans devoted to the classification of lung nodules (see \ref{sec:lungCTChallenge} for further details).

@@

\section{Brain/Kaggle Brain}
Brain cancer is another major type of cancer. According to American Society of Clinical Oncology, "brain and other nervous system cancer is the 10th leading cause of death for men and women. It is estimated that 17760 adults (9910 men and 7850 women) will die from primary cancerous brain and central nervous system tumors this year"~\cite{43}. The following research papers are based on the Kaggle "Brain MRI Images for Brain Tumor Detection" dataset. This dataset, unlike the ones for the other body parts, does not come from a certified medical authority but from the Kaggle website \cite{45}. However, since it is the only dataset available for brain tumor classification, some publications used it. Further details about this dataset can be found in section \ref{sec:kaggleBrain}.

>>>>

\chapter{Literature review}
\label{ch:literature_review}
%Traditionally, building a CAD system consisted of two different steps: "Lesion  detection (CADe) and  false-positive  reduction (CADx).  Lesion  detection  was  primarily  based  on  algorithms  specific  to  the  detection task, resulting in many candidate lesions. False-positive reduction was commonly based on traditional  machine  learning  methods  to  reduce  the  false  positive  rate. However,  even  with  these  complicated and sophisticated programs, the general performance of conventional CAD systems was not good, thus hampering their widespread usage in clinical practice"~\cite{41}.\\
%Nowadays, these two steps tend to disappear thanks to deep learning algorithms, in particular the ones based on convolutional neural networks. In fact, the latter can be considered as single-step CAD system. For this reason, CAD systems are currently a trending topic in deep learning.\\

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Johan}}Since this work focuses on lesions classification, this chapter presents the main research works about CADx systems based on convolutional neural networks for each body part used in sections \ref{ch:paper_reproduction} and \ref{ch:transfer_learning}, i.e. prostate, lung and brain. Furthermore, it exclusively focuses on studies that used the same datasets as the ones used in this work.

\section{Prostate/PROSTATEx}

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}As stated by Gao et al., "Prostate  cancer  is  the  most  common  malignancies  among  men  and  remains  a  second  leading  cause to deaths in men globally. It was predicted that there would be 1.7 million new cancer cases by 2030. The early detection and diagnosis of prostate cancer can help to survive nine out of 10 men for the last five years"\cite{41}. Therefore, researchers proposed many different models to achieve good performance in prostate cancer detection. All the research works mentioned below are based on the SPIE-AAPM-NCI PROSTATEx Challenge dataset. This dataset is composed of multiparametric MRIs (T2W, DWI, ADC, DCE, PD, Ktrans) for a total of 204 training patients and 208 challenge patients (see Section \ref{sec:prostatex_dataset_description}).

@@

\section{Lung/Lung CT Challenge}

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Johan \& Julien}}Gao et al. claimed that "lung cancer is one of the most frequent and leading causes of death all over the world. It was reported that there were approximately $1.8$x$10^6$ new cases of lung cancer globally in 2012. Early detection of lung cancer, which is typically viewed in the form of lung nodules, is an efficient way to improve the survival rate"~\cite{41}. The articles cited below make use of the SPIE-AAPM Lung CT Challenge dataset, which is composed of CT scans devoted to the classification of lung nodules (see \ref{sec:lungCTChallenge} for further details).

@@

\section{Brain/Kaggle Brain}

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Johan}}Brain cancer is another major type of cancer. According to American Society of Clinical Oncology, "brain and other nervous system cancer is the 10th leading cause of death for men and women. It is estimated that 17760 adults (9910 men and 7850 women) will die from primary cancerous brain and central nervous system tumors this year"~\cite{43}. The following research papers are based on the Kaggle "Brain MRI Images for Brain Tumor Detection" dataset. This dataset, unlike the ones for the other body parts, does not come from a certified medical authority but from the Kaggle website \cite{45}. However, since it is the only dataset available for brain tumor classification, some publications used it. Further details about this dataset can be found in section \ref{sec:kaggleBrain}.

### 2020-03-04 Common/chapters/medical.tex
Johan and Julien: all names for each section added

\chapter{Medical information}
\label{ch:medical}
This chapter gives an overview of the basic medical knowledge that is required to apprehend the following chapters smoothly. First of all, some notions about cancer are described in order to understand what it is and which effects it has on the human body. Second, medical data has its own file formats. To make use of them in a deep learning project, medical files must be processed in a certain way, depending on each format. In fact, formats represent 2D, 3D or even 4D data. Some of them require specific normalization in order to get the right rendering. Finally, some visualization tools were developed to display raw medical files easily.

\section{Cancer}
\subsection{Basics}
An accumulation of cells forming a mass is called a tumor. These tumors are detectable thanks to medical imaging (see section \ref{sec:medical_imaging}) and other symptoms. However, not every tumor is as dangerous as the other, as it can be benign (does not contain cancerous cells) or malignant (contains cancerous cells).

The term cancer refers to different phenomena which involve mutation, abnormal multiplication and spreading of cells. As stated by Hanahan et al. in "The Hallmarks of Cancer" \cite{19} and "The Hallmarks of Cancer: The Next Generation" \cite{20}, every malignant tumor acquires six different capabilities during its evolution: 
\begin{itemize}
	\item \textbf{"Sustaining proliferative signaling"}\\ Cancerous cells don't wait for the body's approval to grow and proliferate, contrary to normal cells. They become responsible for their own multiplication.
	\item \textbf{"Evading growth suppressors"}\\
The body sends signals to contain cell growth within a tissue. Cancerous cells are insensitive to these. 
	\item \textbf{"Activating invasion and metastasis"}\\
Metastases are cells whose role is to propagate to other parts of the body in order to colonize and create new tumors. 
	\item \textbf{"Enabling replicative immortality"}\\
Healthy cells replication is limited to a certain amount, which is not the case for cancerous cells. 
	\item \textbf{"Inducing angiogenesis"}
\\Angiogenesis is the process of creating new blood vessels. Tumors have an influence on angiogenesis around them, since they need vascularization to continue growing. 
	\item \textbf{"Resisting cell death"}
\\Apoptosis is the programmed death of cells, which is part of the continuous regeneration of every cell within a body. Cancerous cells survive this programmed death.

\subsection{Seriousness}

Most cancers can be staged thanks to the TNM system. The T corresponds to the tumor size and its location; the N corresponds to whether or not the tumor has spread to draining lymph nodes; the M corresponds to the presence or absence of metastases in other parts of the body \cite{21}. These pieces of information are used to classify cancer between four (I to IV) or sometimes five (0 to IV) different stages, reflecting the progression and the seriousness of the illness \cite{22}. The earlier they are detected, the higher the chances of recovery are. This aspect makes cancer detection critical since every misjudgment can threaten someone's life.

%\subsection{Chronological evolution}

\section{Types of medical imaging}
\label{sec:medical_imaging}

Multiple types of medical imaging exist. The most commonly used to detect cancer are Magnetic Resonance Imaging (MRI), CT (Computed Tomography) scans
%, PET (Positron Emission Tomography) scans
and mammograms.

@@

\section{DICOM file format}
\label{sec:DICOM}

\subsection{Origin}

The acronym DICOM stands for Digital Imaging and Communications in Medicine. Before the 1980’s, images resulting from CT scans and MRIs were only decodable by machine manufacturers, while the medical community needed to export and share them for other tasks. For that reason, the ACR (American College of Radiology) and the NEMA (National Electrical Manufacturers Association) created a committee to build a standard. After two iterations with other names, DICOM was created in 1993. It standardized the representation of medical images and their transmission since it provided a network protocol built on top of TCP/IP.

\subsection{Data format}

DICOM files can be viewed as containers of attributes, also called tags. The values of the pixels themselves are stored under the "Pixel Data" tag. Every single DICOM file usually represents a 2-dimensional image, which will form a 3-dimensional volume when put all together.

@@

\subsection{Processing images}

When manipulating DICOM files, multiple details must be taken into account.

\subsubsection{Order}
First of all, the name of the files within datasets is a 6-digit number, from 000000 to the number of images minus one. However, this order doesn’t match the real order of the images. In fact, the correct order is given by the "Instance Number" tags contained in the various files. Therefore, converted images must be sorted by instance number.

\subsubsection{Data manipulation}
\label{sec:dicom_data_manipulation}
CT and MRI machines, as well as monitors, differ from one manufacturer to the other and even from one model to the other. DICOM takes this problematic into account by providing specific tags that allow to display the exact same representation of the data, no matter the hardware used. Otherwise, physicians may struggle to detect anomalies because of color and exposition-related variations. 
Therefore, before displaying or converting an image to any format (such as png), pixel data must be normalized.

@@

where~$c$ is the window center,~$w$ window width, ~$P$ the pixel input value, ~$y$ the pixel output value, ~$y_{min}$ the minimal value of the output range (usually 0), ~$y_{max}$ the maximal value of the output range (usually 255). Equations \ref{eq:dicom1}, \ref{eq:dicom2} and \ref{eq:dicom3} ensure that the pixel values are correctly distributed within the output range.

\end{enumerate}

\section{NIfTI file format}

\subsection{Origin}
The Neuroimaging Informatics Technology Initiative (NIfTI) file format is the successor of the ANALYZE file format. The main problem of the latter was that it was lacking information about orientation in space. Therefore, the interpretation of stored data could be problematic and inconsistent. For instance, there was a real confusion to determine the left and right sides of brain images. Hence, the NIfTI file format was defined to overcome this major issue.

\subsection{Data format}
Unlike the ANALYZE format that used two files to store the metadata and the actual data, the NIfTI file format stores them in one single file “.nii” but keeps this split between the real data and the header for compatibility. This has the advantage to facilitate the use of the data and avoid storing the data without the meta-data. The NIfTI format can also be compressed/decompressed on-the-fly using the “deflate”  algorithm.

\subsection{Overview of the header structure}
With the goal of preserving the compatibility between the ANALYZE and the NIfTI formats, both headers have the same size of 348 bytes. “Some fields were reused, some were preserved, but ignored, and some were entirely overwritten”. Details about the different fields contained in the header can be found here\footnote{\url{https://brainder.org/2012/09/23/the-nifti-file-format/}}.

\section{RAW and MHD file formats}

Some datasets use a combination of RAW and MHD files. The latter contain metainformation about their corresponding RAW file(s) which contain the data. In most cases, each MHD file points to a unique RAW file whose name is the same as the MHD file name. A single RAW file can be used to represent three-dimensional data, i.e. the combination of multiple two-dimensional images. Libraries such as \mbox{SimpleITK} in Python allow to manipulate RAW images in an easy way.

\section{Visualization tools}
Processing data manually increases the probability of making mistakes. For that matter, visualization tools relying on the same processing code as the ones used to generate training images were developed. Their primary goal is to compare our visual representation of an image to the one obtained in professional pieces of software. These tools are convenient to visualize a dataset easily. In some cases (especially RAW/MHD images), no free software capable of reading the files was available, which made the corresponding tool useful.

\subsection{DICOM}
Visualizing DICOM files is pretty straightforward since each file represents a single two-dimensional image. However, a lot of pixel transformations and normalizations have to be applied to obtain the desired result (see section \ref{sec:dicom_data_manipulation}), which may be a source of errors. Our tools allows to display a single DICOM file as well as a sequence of files if the function is fed with a directory. Users can then scroll through the Z-axis, displaying the next or previous slice.

@@

\subsection{NIfTI}
As a single NIfTI file can contain three- or four-dimensional data (the fourth dimension being time), our visualization tool takes this aspect into consideration by allowing to scroll through them: scrolling with the mouse goes through slices belonging to a specific timestamp over a specific axis, while the left and right arrows allows to jump to the same slice at another timestamp. When reaching the end of a timestamp with the mouse, the first slice of the next timestamp is displayed. Furthermore, the three-dimensionality implies that the volume is viewable from three different perspectives. For example, a three-dimensional brain volume can display it from the top of the head to the bottom, from one ear to the other and from the back of the head to the person's face. Therefore, the user can choose a specific axis to navigate through. If no axis is chosen, all three perspectives are shown one after the other.

@@

\subsection{RAW}
Medical RAW files are three-dimensional, which means that a single file contains a volume, i.e. a succession of 2D slices. Scrolling with the mouse goes through slices over a specific axis. Like the NIfTI file format, the user can profit from the three-dimensionality by displaying the body part from three different perspectives. This tool was particularly useful since no free software capable of handling these files was available. Photography-related programs such as Photoshop can open RAW pictures but not these three-dimensional ones.

@@

\section{Conversion to PNG}

In addition to the visualization tools, libraries allowing to convert medical files into PNG files were developped. The main one is called "anydatasettopng" and is the one to use to convert any file format to PNG since it makes use of the others. Given a root directory, it scans the latter and its subdirectories to find all DICOM, NIfTI and RAW/MHD files. Then, each file is converted to PNG and saved into the same directory as the one it came from. This library allows to keep an easily accessible visual representation of medical files without having to rely on a specific piece of software to display them. Finally, it also played the role of debugging tool since the conversion techniques are the same as the ones used further in this work.

\subsection{8-bit conversion}

Medical data is often represented over 16 bits. However, exporting images to PNG require 8-bit images. To transpose a 16-bit image to an 8-bit one, the procedure described by algorithm \ref{alg:16_to_8_bits_conversion} was applied.

>>>>

\chapter{Medical information}
\label{ch:medical}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}This chapter gives an overview of the basic medical knowledge that is required to apprehend the following chapters smoothly. First of all, some notions about cancer are described in order to understand what it is and which effects it has on the human body. Second, medical data has its own file formats. To make use of them in a deep learning project, medical files must be processed in a certain way, depending on each format. In fact, formats represent 2D, 3D or even 4D data. Some of them require specific normalization in order to get the right rendering. Finally, some visualization tools were developed to display raw medical files easily.

\section{Cancer}
\subsection{Basics}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}An accumulation of cells forming a mass is called a tumor. These tumors are detectable thanks to medical imaging (see section \ref{sec:medical_imaging}) and other symptoms. However, not every tumor is as dangerous as the other, as it can be benign (does not contain cancerous cells) or malignant (contains cancerous cells).

The term cancer refers to different phenomena which involve mutation, abnormal multiplication and spreading of cells. As stated by Hanahan et al. in "The Hallmarks of Cancer" \cite{19} and "The Hallmarks of Cancer: The Next Generation" \cite{20}, every malignant tumor acquires six different capabilities during its evolution: 
\begin{itemize}
	\item \textbf{"Sustaining proliferative signaling"}\\ Cancerous cells don't wait for the body's approval to grow and proliferate, contrary to normal cells. They become responsible for their own multiplication.
	\item \textbf{"Evading growth suppressors"}\\
The body sends signals to contain cell growth within a tissue. Cancerous cells are insensitive to these. 
	\item \textbf{"Activating invasion and metastasis"}\\
Metastases are cells whose role is to propagate to other parts of the body in order to colonize and create new tumors. 
	\item \textbf{"Enabling replicative immortality"}\\
Healthy cells replication is limited to a certain amount, which is not the case for cancerous cells. 
	\item \textbf{"Inducing angiogenesis"}
\\Angiogenesis is the process of creating new blood vessels. Tumors have an influence on angiogenesis around them, since they need vascularization to continue growing. 
	\item \textbf{"Resisting cell death"}
\\Apoptosis is the programmed death of cells, which is part of the continuous regeneration of every cell within a body. Cancerous cells survive this programmed death. 
\end{itemize}

\subsection{Seriousness}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Most cancers can be staged thanks to the TNM system. The T corresponds to the tumor size and its location; the N corresponds to whether or not the tumor has spread to draining lymph nodes; the M corresponds to the presence or absence of metastases in other parts of the body \cite{21}. These pieces of information are used to classify cancer between four (I to IV) or sometimes five (0 to IV) different stages, reflecting the progression and the seriousness of the illness \cite{22}. The earlier they are detected, the higher the chances of recovery are. This aspect makes cancer detection critical since every misjudgment can threaten someone's life.

\section{Types of medical imaging}
\label{sec:medical_imaging}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Multiple types of medical imaging exist. The most commonly used to detect cancer are Magnetic Resonance Imaging (MRI), CT (Computed Tomography) scans
%, PET (Positron Emission Tomography) scans
and mammograms.

@@

\section{DICOM file format}
\label{sec:DICOM}
\subsection{Origin}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}The acronym DICOM stands for Digital Imaging and Communications in Medicine. Before the 1980’s, images resulting from CT scans and MRIs were only decodable by machine manufacturers, while the medical community needed to export and share them for other tasks. For that reason, the ACR (American College of Radiology) and the NEMA (National Electrical Manufacturers Association) created a committee to build a standard. After two iterations with other names, DICOM was created in 1993. It standardized the representation of medical images and their transmission since it provided a network protocol built on top of TCP/IP.

\subsection{Data format}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}DICOM files can be viewed as containers of attributes, also called tags. The values of the pixels themselves are stored under the "Pixel Data" tag. Every single DICOM file usually represents a 2-dimensional image, which will form a 3-dimensional volume when put all together.

@@

\subsection{Processing images}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}When manipulating DICOM files, multiple details must be taken into account.

\subsubsection{Order}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}First of all, the name of the files within datasets is a 6-digit number, from 000000 to the number of images minus one. However, this order doesn’t match the real order of the images. In fact, the correct order is given by the "Instance Number" tags contained in the various files. Therefore, converted images must be sorted by instance number.

\subsubsection{Data manipulation}
\label{sec:dicom_data_manipulation}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}CT and MRI machines, as well as monitors, differ from one manufacturer to the other and even from one model to the other. DICOM takes this problematic into account by providing specific tags that allow to display the exact same representation of the data, no matter the hardware used. Otherwise, physicians may struggle to detect anomalies because of color and exposition-related variations. 
Therefore, before displaying or converting an image to any format (such as png), pixel data must be normalized.

@@

where~$c$ is the window center,~$w$ window width, ~$P$ the pixel input value, ~$y$ the pixel output value, ~$y_{min}$ the minimal value of the output range (usually 0), ~$y_{max}$ the maximal value of the output range (usually 255). Equations \ref{eq:dicom1}, \ref{eq:dicom2} and \ref{eq:dicom3} ensure that the pixel values are correctly distributed within the output range. 
\end{enumerate}

\section{NIfTI file format}
\subsection{Origin}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}The Neuroimaging Informatics Technology Initiative (NIfTI) file format is the successor of the ANALYZE file format. The main problem of the latter was that it was lacking information about orientation in space. Therefore, the interpretation of stored data could be problematic and inconsistent. For instance, there was a real confusion to determine the left and right sides of brain images. Hence, the NIfTI file format was defined to overcome this major issue.

\subsection{Data format}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Unlike the ANALYZE format that used two files to store the metadata and the actual data, the NIfTI file format stores them in one single file “.nii” but keeps this split between the real data and the header for compatibility. This has the advantage to facilitate the use of the data and avoid storing the data without the meta-data. The NIfTI format can also be compressed/decompressed on-the-fly using the “deflate”  algorithm.

\subsection{Overview of the header structure}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}With the goal of preserving the compatibility between the ANALYZE and the NIfTI formats, both headers have the same size of 348 bytes. “Some fields were reused, some were preserved, but ignored, and some were entirely overwritten”. Details about the different fields contained in the header can be found here\footnote{\url{https://brainder.org/2012/09/23/the-nifti-file-format/}}.

\section{RAW and MHD file formats}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Some datasets use a combination of RAW and MHD files. The latter contain metainformation about their corresponding RAW file(s) which contain the data. In most cases, each MHD file points to a unique RAW file whose name is the same as the MHD file name. A single RAW file can be used to represent three-dimensional data, i.e. the combination of multiple two-dimensional images. Libraries such as \mbox{SimpleITK} in Python allow to manipulate RAW images in an easy way.

\section{Visualization tools}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}
Processing data manually increases the probability of making mistakes. For that matter, visualization tools relying on the same processing code as the ones used to generate training images were developed. Their primary goal is to compare our visual representation of an image to the one obtained in professional pieces of software. These tools are convenient to visualize a dataset easily. In some cases (especially RAW/MHD images), no free software capable of reading the files was available, which made the corresponding tool useful.

\subsection{DICOM}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Visualizing DICOM files is pretty straightforward since each file represents a single two-dimensional image. However, a lot of pixel transformations and normalizations have to be applied to obtain the desired result (see section \ref{sec:dicom_data_manipulation}), which may be a source of errors. Our tools allows to display a single DICOM file as well as a sequence of files if the function is fed with a directory. Users can then scroll through the Z-axis, displaying the next or previous slice.

@@

\subsection{NIfTI}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}As a single NIfTI file can contain three- or four-dimensional data (the fourth dimension being time), our visualization tool takes this aspect into consideration by allowing to scroll through them: scrolling with the mouse goes through slices belonging to a specific timestamp over a specific axis, while the left and right arrows allows to jump to the same slice at another timestamp. When reaching the end of a timestamp with the mouse, the first slice of the next timestamp is displayed. Furthermore, the three-dimensionality implies that the volume is viewable from three different perspectives. For example, a three-dimensional brain volume can display it from the top of the head to the bottom, from one ear to the other and from the back of the head to the person's face. Therefore, the user can choose a specific axis to navigate through. If no axis is chosen, all three perspectives are shown one after the other.

@@

\subsection{RAW}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Medical RAW files are three-dimensional, which means that a single file contains a volume, i.e. a succession of 2D slices. Scrolling with the mouse goes through slices over a specific axis. Like the NIfTI file format, the user can profit from the three-dimensionality by displaying the body part from three different perspectives. This tool was particularly useful since no free software capable of handling these files was available. Photography-related programs such as Photoshop can open RAW pictures but not these three-dimensional ones.

@@

\section{Conversion to PNG}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}In addition to the visualization tools, libraries allowing to convert medical files into PNG files were developped. The main one is called "anydatasettopng" and is the one to use to convert any file format to PNG since it makes use of the others. Given a root directory, it scans the latter and its subdirectories to find all DICOM, NIfTI and RAW/MHD files. Then, each file is converted to PNG and saved into the same directory as the one it came from. This library allows to keep an easily accessible visual representation of medical files without having to rely on a specific piece of software to display them. Finally, it also played the role of debugging tool since the conversion techniques are the same as the ones used further in this work.

\subsection{8-bit conversion}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Medical data is often represented over 16 bits. However, exporting images to PNG require 8-bit images. To transpose a 16-bit image to an 8-bit one, the procedure described by algorithm \ref{alg:16_to_8_bits_conversion} was applied.
